Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete

### 2019-05-15 analysis/paper/paper.Rmd
Cut downs and edits

```{r boxplot,  fig.dim=c(9,5), dpi=300,}

ggplot(tombs,aes(total_finds))+geom_histogram()+xlim(0,200)

@@

ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
      theme(legend.title=element_blank(),
        text=element_text(family="Garamond", size=12),
        axis.text.x = element_text(angle = 70, hjust = 1),
        panel.grid.major.x = element_blank())+
            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
               scale_fill_brewer(palette = "Set3")+
                scale_x_reverse()+
                  scale_y_continuous()+
                  facet_zoom(x=variable<900& variable>640,y=value<20)

ggplot(aorist_melt,aes(variable,value,fill=period))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
      theme(legend.title=element_blank(),
        text=element_text(family="Garamond", size=12),
        axis.text.x = element_text(angle = 70, hjust = 1),
        panel.grid.major.x = element_blank())+
            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
               scale_fill_brewer(palette = "Set3")+
                scale_x_reverse()+
                  scale_y_continuous(limits=c(0,20))+geom_label(data=subset(aorist_melt,value>19),aes(label=tomb))

```{r dpi=300}
tombs_gini<-matrix(0, ncol = 0, nrow = 1)
tombs_gini<-data.frame(tombs_gini)
for(i in colnames(aorist_tomb_values[2:51])) {
  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
}

@@

ggplot(tombs_gini,aes(decade,gini_coefficient))+geom_line()+
      xlim(1100,600)+ylim(0.3,0.8)+theme_minimal()+
        theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Date (Years BC)", y="Gini Coefficient")+
                 scale_fill_brewer(palette = "Set2")

@@

plot(Lc(aorist_tomb_values$`800`))

```{r aorist_burials}

aorist_pithoi<-colSums(pithoi[25:74])
aorist_pithoi<-data.frame(aorist_pithoi)

colnames(aorist_pithoi)<-c("aoristic_sum")
row.names(aorist_pithoi)<-dates
aorist_pithoi<-data.frame(aorist_pithoi)
setDT(aorist_pithoi, keep.rownames = "decade")
aorist_pithoi$decade<-as.numeric(aorist_pithoi$decade)

aorist_all_urns<-colSums(all_urns[25:74])
aorist_all_urns<-data.frame(aorist_all_urns)

colnames(aorist_all_urns)<-c("aoristic_sum")
row.names(aorist_all_urns)<-dates
aorist_all_urns<-data.frame(aorist_all_urns)
setDT(aorist_all_urns, keep.rownames = "decade")
aorist_all_urns$decade<-as.numeric(aorist_all_urns$decade)

ggplot(NULL)+geom_col(data=aorist_all_urns,aes(decade,aoristic_sum,fill="All urns"))+
  geom_col(data=aorist_pithoi,aes(decade,aoristic_sum,fill="Pithoi"))+
      xlim(1100,630)+theme_minimal()+
        theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Date (Years BC)", y="Aoristic Density")+
                 scale_fill_brewer(palette = "Set2")

```{r aorist_tomb_adjust}

no_tombs<-colSums(tombs[158:207]!=0)
no_tombs<-data.frame(no_tombs)
colnames(no_tombs)<-c("no_tombs")
row.names(no_tombs)<-dates
no_tombs<-data.frame(no_tombs)
setDT(no_tombs, keep.rownames = "decade")
no_tombs$decade<-as.numeric(no_tombs$decade)

aorist_all_tombs1<-aorist_all_tombs
aorist_by_tomb<-merge(aorist_all_tombs1,no_tombs)

aorist_by_tomb$aorist_by_tomb<-aorist_by_tomb$aoristic_sum/aorist_by_tomb$no_tombs

ggplot(aorist_by_tomb,aes(decade,aorist_by_tomb))+geom_col()+xlim(1100,600)

#Now for the 9th century tombs

tombs_9th<-subset(tombs,tombs$century_built=="9th")
no_tombs_9th<-colSums(tombs_9th[158:207]!=0)

View(tombs_9th)
no_tombs_9th<-data.frame(no_tombs_9th)
colnames(no_tombs_9th)<-c("no_tombs")
row.names(no_tombs_9th)<-dates
no_tombs_9th<-data.frame(no_tombs_9th)
setDT(no_tombs_9th, keep.rownames = "decade")
no_tombs_9th$decade<-as.numeric(no_tombs_9th$decade)

aorist_tombs_9th<-colSums(tombs_9th[158:207])
aorist_tombs_9th<-data.frame(aorist_tombs_9th)
colnames(aorist_tombs_9th)<-c("aoristic_sum")
row.names(aorist_tombs_9th)<-dates
aorist_tombs_9th<-data.frame(aorist_tombs_9th)
setDT(aorist_tombs_9th, keep.rownames = "decade")
aorist_tombs_9th$decade<-as.numeric(aorist_tombs_9th$decade)

aorist_by_tomb_9th<-merge(aorist_tombs_9th,no_tombs_9th)
aorist_by_tomb_9th$aorist_by_tomb<-aorist_by_tomb_9th$aoristic_sum/aorist_by_tomb_9th$no_tombs

ggplot()+geom_col(data=aorist_by_tomb_9th,aes(decade,aorist_by_tomb,fill="9th Century Tombs"))+
  geom_col(data=aorist_by_tomb,aes(decade,aorist_by_tomb, fill="All Tombs"),alpha=0.7)+
    xlim(1100,600)+theme_minimal()+
        theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Date (Years BC)", y="Aoristic Density/Number of Tombs Represented")+
                 scale_fill_brewer(palette = "Set2")

names(tombs)
tombs_in_use<-colSums(tombs[6:15]!="")
tombs_in_use<-data.frame(tombs_in_use)
colnames(tombs_in_use)<-c("no_tombs")
tombs_in_use<-data.frame(tombs_in_use)
setDT(tombs_in_use, keep.rownames = "period")
tombs_in_use$period<-factor(tombs_in_use$period,levels=c("SM", "EPG", "MPG", "LPG", "PGB", "EG", "MG", "LG", "EO", "LO"))

ggplot(no_tombs_9th,aes(decade,no_tombs, fill="Tombs in Use"))+geom_col()+theme_classic()+
     theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
            xlim(1100,600)+
              labs(x="Date (Years BC)", y="Number of Tombs in Use")+
                 scale_fill_brewer(palette = "Set2")

ggplot(tombs_in_use,aes(period,no_tombs, fill="Tombs in Use"))+geom_col()+theme_classic()+
        theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Period", y="Number of Tombs in Use")+
                 scale_fill_brewer(palette = "Set2")

ggplot(tombs,aes(start_period, fill=tomb_type_condensed))+geom_bar()+theme_classic()+
        theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Period", y="Number of New Tombs")+
                 scale_fill_brewer(palette = "Set2")

```{r rates of burials}

tombs_ranked<-tombs[with(tombs, order(-tombs$total_pottery)), ]
tombs_top_12<-head(tombs_ranked,12)

View(tombs_top_12)

no_tombs_top<-colSums(tombs_top_12[158:207]!=0)

no_tombs_top<-data.frame(no_tombs_top)
colnames(no_tombs_top)<-c("no_tombs")
row.names(no_tombs_top)<-dates
no_tombs_top<-data.frame(no_tombs_top)
setDT(no_tombs_top, keep.rownames = "decade")
no_tombs_top$decade<-as.numeric(no_tombs_top$decade)

aorist_tombs_top<-colSums(tombs_top_12[158:207])
aorist_tombs_top<-data.frame(aorist_tombs_top)
colnames(aorist_tombs_top)<-c("aoristic_sum")
row.names(aorist_tombs_top)<-dates
aorist_tombs_top<-data.frame(aorist_tombs_top)
setDT(aorist_tombs_top, keep.rownames = "decade")
aorist_tombs_top$decade<-as.numeric(aorist_tombs_top$decade)

aorist_by_tomb_top<-merge(aorist_tombs_top,no_tombs_top)
aorist_by_tomb_top$aorist_by_tomb<-aorist_by_tomb_top$aoristic_sum/aorist_by_tomb_top$no_tombs

ggplot()+geom_col(data=aorist_by_tomb_top,aes(decade,aorist_by_tomb,fill="12 Largest Tombs"))+
  geom_col(data=aorist_by_tomb,aes(decade,aorist_by_tomb, fill="All Tombs"),alpha=0.7)+
    xlim(1100,600)+theme_minimal()+
        theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Date (Years BC)", y="Aoristic Density/Number of Tombs Represented")+
                 scale_fill_brewer(palette = "Set2")

```{r top_tombs, fig.dim = c(7, 5), dpi=300, fig.align='center',}

cols<-c(1,158:207)
aorist_all_tomb_names<-tombs_top_12[,cols]
aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
colnames(aorist_all_tomb_names)<-c("tomb",dates)

aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
str(aorist_all_tomb_melt)
aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))

top_12<-tombs_top_12$tomb
aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_12)

ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb)+
  geom_col()+theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="Garamond", size=12))+
          labs(x="Date (Years BC)", y="Aoristic Sum")+
              scale_fill_brewer(palette = "Set3")+ 
                theme(legend.position="none")+
                  coord_cartesian(ylim=c(0,30),xlim=c(1100,610))+scale_y_continuous(expand=c(0,0))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

>>>>

Also apparent from Table 1, though, are the markedly low rates of burial which edominated at both cemeteries. Two to four burials per tomb per century amounts to a very selective burial rite, even considering the restricted portion of the population probably represented by these cemeteries [@Cavanagh1996, 664]. It seems unlikely, then, that the rising rates of burial noted above are, for instance, the direct result of population growth; the increase seems too rapid, and is concentrated only in certain (odler) tombs, not to mention that, based on KULP’s survey data, the major population increase at EIA Knossos should be dated a century or more before the late 8th century peak in burial, in the Protogeometric period (Whitelaw, per comm.).

We might assume that the tombs which became the focus of intense burial activity in the late 8th and early 7th centuries would be those which yielded the most burials and grave goods when excavated. Certainly, the largest 12 tombs are all chamber tombs, with 10 dating to the 9th century, and two to the 10th. But looking more closely at the aoristic sums of their assemblages through time, not all of them appear to have exhibited the noted intensification of burial activity into the Orientalising period.

```{r top_tombs, fig.dim = c(7, 5), dpi=300, fig.align='center',fig.cap="Aoristic sums for the 12 largest tombs (in terms of grave goods), arranged in decreasing order of total recovered artefacts. Note, for the sake of better illustrating the relative patterns, rather than absolute values, the scales on each plot are distinct."}

cols<-c(1,158:207)
aorist_all_tomb_names<-tombs_top_20[,cols]
aorist_all_tomb_names<-data.frame(aorist_all_tomb_names)
colnames(aorist_all_tomb_names)<-c("tomb",dates)

aorist_all_tomb_melt<-melt(aorist_all_tomb_names)
aorist_all_tomb_melt$variable<-as.numeric(as.character(aorist_all_tomb_melt$variable))

top_20<-tombs_top_20$tomb
aorist_all_tomb_melt$tomb<-factor(aorist_all_tomb_melt$tomb,levels=top_20)

ggplot(aorist_all_tomb_melt,aes(variable, value))+geom_col()+theme(legend.position = "none")+facet_wrap(~tomb,scales="free_y")+
  geom_col()+theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="Garamond", size=12))+
          labs(x="Date (Years BC)", y="Aoristic Sums")+
              scale_fill_brewer(palette = "Set3")+ 
                theme(legend.position="none")+
                  coord_cartesian(xlim=c(1100,610))+scale_y_continuous()+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,610))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

From Figure \@ref(fig:top_tombs), we can see that, though seven of the 12 did exhibit late, intense episodes of deposition, tombs X, G, Q, 104 and 283 evince a different pattern, with their respective peaks coming in the PGB and EG periods, around a century earlier. We have seen already in Figure \@ref(fig:aoristbycem) that, particularly in the KNC, the end of the 9th century was another period of increased burial activity

```{r boxplot,  fig.dim=c(7,4), dpi=300,}

tombs_top_20<-head(tombs_ranked,20)
top20<-as.vector(tombs_top_20$tomb)
pots_top_20$tomb<-factor(pots_top_20$tomb,levels=top20)

pots_top_20<-pots[is.element(pots$tomb, top20),]
pots_top_20_9th_7th<-subset(pots_top_20,century=="9th"|century=="8th"|century=="7th")
ggplot(pots_top_20_9th_7th,aes(tomb,fill=tomb_century))+geom_bar()+facet_wrap(~century)

tombs$group<-ifelse(tombs$tomb=="X"|tombs$tomb=="G"|tombs$tomb=="Q"|tombs$tomb=="104"|tombs$tomb=="283","A",
                   ifelse(tombs$tomb=="P"|tombs$tomb=="292"|tombs$tomb=="75"|tombs$tomb=="107"|tombs$tomb=="285"|tombs$tomb=="II"|tombs$tomb=="218","B","OTHER"))
names(tombs)
ggplot(tombs,aes(antique,fill=group))+geom_histogram()

@@

ggplot(aorist_melt,aes(variable,value,fill=period,outlier.size=0.3))+geom_boxplot(aes(group = cut_width(variable, 10)))+theme_bw()+
      theme(legend.title=element_blank(),
        text=element_text(family="Garamond", size=12),
        axis.text.x = element_text(angle = 70, hjust = 1),
        panel.grid.major.x = element_blank())+
            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
               scale_fill_brewer(palette = "Set3")+
                scale_x_reverse(expand=c(0,0))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,20))
names(tombs)
tombs$century_built<-factor(tombs$century_built,levels=c("11th","10th","9th","8th","7th"))
ggplot(tombs,aes(start_period,dromos_length))+geom_boxplot()
ggplot(data=subset(pots,pots$import=="Y"),aes(period))+geom_histogram()+xlim(1100,630)
```

```{r dpi=300}
aorist_tomb_values<-tombs[c(1,158:207)]
aorist_tomb_values[aorist_tomb_values == 0] <- NA

tombs_gini<-matrix(0, ncol = 0, nrow = 1)
tombs_gini<-data.frame(tombs_gini)
for(i in colnames(aorist_tomb_values[2:51])) {
  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
}

@@

@@
